PLAE enables quick and easy searching across one million ocular transcriptomes

### 2022-06-17 manuscript_00.Rmd
tweaks

We used a "pseudo bulk" based approach wherein the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping. The creates a counts matrix which has statistical properties matching that of "bulk" RNA seq experiment matrices. This approach has been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021]. Once the aggregated matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes. We tested each species (Human, Macaque, Mouse) separately. We had an insufficient number of chicken studies to conduct differential testing. The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential expression. After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g. cones versus all remaining cell types) or in a pair-wise manner (cone against rod). We fin
```{r}
# fix  occasional NA
# data made from src/ensdb.R
diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')

@@

Many of these genes can be differentially expressed in more than one cell type. To identify genes that are consistently differentially expressed we can use a more specific test. We define a consistent gene that is

```{r}
scEiaD_2020_v01 %>% 
    tbl("diff_testing") %>% 
    filter(Base == 'Cone') %>%  
    filter(Group == 'CellType (Predict)', 
           Against == 'All') %>% 
  group_by(Gene) %>% 
  summarise(Count = n(), log2FoldChange = mean(log2FoldChange), padj_count = sum(padj < 1e-4), padj = mean(padj)) %>% 
  filter(!is.na(log2FoldChange), padj_count > 1) %>%  
  arrange(-log2FoldChange)

>>>>

We used a "pseudo bulk" based approach wherein the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping. The creates a counts matrix which has statistical properties matching that of "bulk" RNA seq experiment matrices. This approach has been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021]. Once the aggregated matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes. We tested each species (Human, Macaque, Mouse) separately. We had an insufficient number of chicken studies to conduct differential testing. The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential expression. After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g. cones versus all remaining cell types) or in a pair-wise manner (cone against rod). We find
```{r}
# fix  occasional NA
# data made from src/ensdb.R
diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')

@@

# Well supported ocular cell types

Cell types with more than two independent datasets in both human and mouse.
```{r}
ws_oct <- meta_filter %>%  
    filter(Organ == 'Eye') %>% 
    group_by(organism, CellType_predict, study_accession) %>% 
    count() %>% 
    filter(n >= 50) %>% 
    ungroup() %>% 
    group_by(organism, CellType_predict) %>% 
    count() %>% filter(n >= 3) %>% 
  ungroup() %>% 
  group_by(CellType_predict) %>% 
  summarise(Count = n()) %>% 
  filter(Count > 1, !is.na(CellType_predict)) %>% 
  pull(CellType_predict)
```

To identify genes that are differentially expressed across species we

```{r}
consist_diff <- scEiaD_2020_v01 %>% 
    tbl("diff_testing") %>% 
    filter(Base %in% ws_oct) %>%  
    filter(Group == 'CellType (Predict)', 
           Against == 'All') %>% 
    group_by(Base, Gene) %>% 
    summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-4), `mean padj` = mean(padj)) %>% 
    filter(`mean log2FC` > 2, `sig count` > 1, `mean padj` < 1e-5) %>%  
    collect() %>% 
    mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
    left_join(gene_tab, by = 'GENEID') %>% 
    select(-Gene) %>% 
    relocate(SYMBOL, GENEID) %>% 
    dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
    arrange(-`mean log2FC`)
